---
title: "Cause of Death"
author: "Teresa Chen"
date: "2/24/2019"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(rio)
library(here)
library(gganimate)
library("paletteer")

knitr::opts_chunk$set(echo = TRUE)
```



```{r importdata, include=FALSE}

# Import dataframe and rename it as df

df <- import(here::here("data", "Leading_Causes_of_Death__United_States.csv"),
              setclass = "tbl_df") %>% 
  janitor::clean_names()


```


# Overview of data

```{r tidydf}

#str(df)

# clean dataframe to keep useful column  

tidy_df <- df %>% 
  dplyr::select(year, cause_name, state, deaths, age_adjusted_death_rate) %>%   
  mutate(cause = as.factor(cause_name),
         state = as.factor(state),
         rate = age_adjusted_death_rate/100) %>%  #rate is in decimal value 
  dplyr::select(year, state, cause, deaths, rate)

head(tidy_df)
skimr::skim(tidy_df)
```

# Summary of data:

1. No missing data. 
2. There are `r length(unique(tidy_df$cause))` cuases of death, including 
`r unique(tidy_df$cause)`
3. There are `r length(unique(tidy_df$state))` states. 
4. Dataframe has data from 1999 to 2016 (`r 2016-1999+1` years).


# Plot 1

The first plot is intended to deliver a message to **general population** and is trying to answer *what leading death causes are across years*. 

```{r plot1_v1, fig.width=15, fig.height=20}




plot1_v1 <- tidy_df %>% 
  # filter out "all causes"
  filter(cause != "All causes") %>% 
  
  # replace "CLRD" with its whole name
  mutate(cause = replace(as.character(cause), 
                         cause == "CLRD", 
                         "Chronic lower respiratory diseases")) %>%
  
  # calculate yearly total deaths from all states
  group_by(year, cause) %>% 
  mutate(deaths_by_year_cause = sum(deaths)) %>%
  ungroup() %>% 
  
  # plot
  ggplot(aes(x = fct_reorder(cause, deaths_by_year_cause), 
             y = deaths_by_year_cause/1000000)) +
    geom_col(fill = "steelblue") +
    scale_y_continuous(expand = c(0, 0)) +
    coord_flip() +
    facet_wrap(~year, ncol = 3) +
    theme_minimal(base_size = 20) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold")) +
    labs(y = "No. of deaths (in millions)",
         x = "Causes of deaths",
         title = "Heart disease is a leading cause of death over years")
         

plot1_v1
```

## What can be improved?

1. Too many facets. A total of 18 facets make it hard to compare causes of death across years.
2. Rank of causes changes. `fct_reorder` uses cannot reflect an accurate rank in each year (ex. 2016)

# Plot 1: version 2

```{r plot1_v2}

plot1_v2 <- tidy_df %>% 
  filter(cause != "All causes") %>% 
  mutate(cause = replace(as.character(cause), 
                         cause == "CLRD", 
                         "Chronic lower respiratory diseases")) %>% 
  group_by(year, cause) %>% 
  mutate(deaths_by_year_cause = sum(deaths)) %>%
  ungroup() %>% 
  
  # plot
  ggplot(aes(x = fct_reorder(cause, deaths_by_year_cause), 
             y = deaths_by_year_cause/1000000)) +
    geom_col(fill = "steelblue") +
    scale_y_continuous(expand = c(0, 0)) +
    coord_flip() +
    theme_minimal(base_size = 15) +
    theme(panel.grid.minor = element_blank(),
          plot.margin = margin(1, 0.5, 1, 0.5, "cm")) +
    transition_time(year) +
    labs(title = "Top 10 Causes of Deaths",
         subtitle = "Year: {round(frame_time)}",
         caption = "Source: Centers for Disease Control and Prevention",
         y = "No. of deaths (in millions)",
         x = "Causes of deaths")


animate(plot1_v2, duration = 25,  
                  nframes = 100, 
                  end_pause = 10,
                  renderer = gifski_renderer(width = 1000))

```



```{r plot1_v3}

plot1_v3 <- tidy_df %>% 
  filter(cause != "All causes") %>% 
  mutate(cause = replace(as.character(cause), 
                         cause == "CLRD", 
                         "Chronic lower respiratory diseases")) %>%
  group_by(year, cause) %>% 
  mutate(deaths_by_year_cause = sum(deaths)) %>%
  
  #choose one state to rank the deaths_by_year_cause (rank 1-10) each year
  filter(state =="Oregon") %>% 
  group_by(year) %>% 
  arrange(desc(deaths_by_year_cause)) %>% 
  mutate(rank = row_number()) %>% 
  arrange(year) %>% 
  ungroup() %>% 
  
  #plot
  ggplot(aes(x = '{frame_time}', y = rank)) +
  geom_text(aes(label = cause, 
                color = cause,
                group = cause),
            size = 8) +
  scale_x_discrete(labels = NULL) +
  scale_y_reverse() +
  scale_color_paletteer_d(rcartocolor, Vivid) +
  transition_states(year, 
                    transition_length = 1,
                    state_length = 3) +
  ease_aes('sine-in-out') +
  theme_void(base_size = 30) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = "Rank of causes of death: {closest_state}") +
  guides(color = "none")
  
  

animate(plot1_v3, duration = 30, nframes = 300,
        renderer = gifski_renderer(width = 1200))
```




